The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2019-10-23 categorical-zero-one.tex
some more introductory stuff

Throughout, $\cC$ is a Markov category in the sense of \cite[Definition~2.1]{markov_cats} (see also the closely related earlier work of Golubtsov~\cite{golubtsov} and Cho and Jacobs~\cite{cho_jacobs}). $J$ is a possibly infinite set, used for indexing products of objects, and $F \subset J$ ranges over finite subsets.

@@

@@

\section{Infinite tensor products in semicartesian monoidal categories}

Suppose $\cC$ is a semicartesian monoidal category.
To make sense of infinite families of random variables within $\cC$,
we will want to have a notion of infinite tensor product in $\cC$. By Kolmogorov's extension theorem, the joint distribution of infinitely many random variables is uniquely determined by the family of distributions of all finite subsets of these variables. This motivates a general definition of the infinite tensor product of a family of objects $(X_i)_{i \in J}$, based on the observation that if $F \subseteq F' \subseteq J$ are two finite subset, then the fact that $\cC$ is semicartesian monoidal gives us morphisms
\[
	\bigotimes_{i \in F'} X_i \longrightarrow \bigotimes_{i \in F} X_i
\]
to be interpreted as \emph{marginalization}. Via these maps, the finite tensor products $\bigotimes_{i \in F} X_i$ make up a cofiltered diagram in the form of a functor from the poset of finite subsets of $J$, ordered by reverse inclusion, to $\cC$.

@@

Of course, in a general Markov category the above cofiltered limit need not exist for infinite $J$, in which case the infinite tensor product does not exist either (as in \Cref{infprods_finstoch}).
\section{Infinite tensor products in Markov categories}

@@

\todo[inline]{Is this good terminology?}

@@

\begin{remark}
    Condition (2) is also equivalent to the statement that, given a compatible family of deterministic maps $a_F: A \to \bigotimes_{i \in F}$, the induced map $a: A \to \bigotimes_i X_i$ is deterministic.
\end{remark}

>>>>

For monoidal categories, we make frequent use of string diagram notation. We omit object labels whenever these are obvious from the context.

In \Cref{infprod_semicartesian}, $\cC$ denotes a semicartesian monoidal category. In \Cref{infprod_markov} and after, $\cC$ denotes a Markov category in the sense of \Cref{markov_cat}. $J$ is a set used for indexing products of objects; the definitions and results of this paper are nontrivial only when $J$ is infinite. $F$ denotes either an arbitrary finite set or more concretely a finite subset of $J$.

@@

A monoidal category $\cC$ is \emph{semicartesian} if the monoidal unit $I \in \cC$ is terminal. Equivalently, it is a monoidal category equipped with morphisms
\[
	X \otimes Y \longrightarrow X, \qquad X \otimes Y \longrightarrow Y,
\]
which are natural in $X$ and $Y$ and coincide with the monoidal structure isomorphisms whenever $X = I$ or $Y = I$. Here are the main examples that we will be considering.

\begin{example}
	$\FinStoch$ is the category of finite sets with \emph{stochastic maps}, where a stochastic map $f : X \to Y$ is a matrix $(f_{xy})_{x \in X,y \in Y}$ of nonnegative real numbers such that $\sum_y f_{xy} = 1$ for every $x$. Stochastic maps compose by matrix multiplication, which in this context is also known as the \emph{Chapman--Kolmogorov equation}. We consider $\FinStoch$ as a monoidal category with respect to the cartesian product of sets on objects, and as one would expect the Kronecker product of stochastic maps on morphisms.
\end{example}

\begin{example}
	Let $\CRing_+$ be the category of commutative rings with maps which are merely additive and unit-preserving, considered as a monoidal category in the obvious way with respect to the tensor product of rings. Then since $\CRing_+$ has the monoidal unit $\Z$ as its initial object, we can conclude that $\CRing_+\op$ is semicartesian monoidal.
\end{example}

\begin{example}
	$\Stoch$ is the category of measurable spaces $(X,\Sigma_X)$ as objects and Markov kernels as morphisms. We sketch the definition here and refer to~\cite[Section~4]{markov_cats} and references therein for the details. A \emph{Markov kernel} $(X,\Sigma_X) \to (Y,\Sigma_Y)$ is a function
	\[
		f \: : \: \Sigma_Y \times X \longrightarrow [0,1], \quad (S,x) \longmapsto f(S|x)
	\]
	assigning to every $x \in X$ a probability measure $f(-|x) : \Sigma_Y \to [0,1]$ in such a way that for every $S \in \Sigma_Y$, the function $x \mapsto f(S|x)$ is measurable. Intuitively, $f$ assigns to every $x \in X$ a random element of $Y$. Composition is again defined by a variant of the Chapman--Kolmogorov equation, and the monoidal structure is defined in terms of the usual product of measurable spaces.
\end{example}

In the graphical calculus, the unique morphism $X \to I$ for an object $X \in \cC$ is denoted by
\[
	\tikzfig{terminal}
\]
The projection maps $X \otimes Y \to X$ and $X \otimes Y \to Y$ are correspondingly written as
\[
	\tikzfig{marginals}
\]

Our goal is to use semicartesian monoidal categories such as the above in order to develop aspects of probability theory in categorical terms. As it turns out, doing so requires a bit more structure, in a form which has been axiomatized in the notion of \emph{Markov category} \cite[Definition~2.1]{markov_cats}.

\begin{definition}
	A Markov category $\cC$ is a semicartesian monoidal category \ldots
	\label{markov_cat}
\end{definition}

A closely related definition was also used in earlier work of Golubtsov~\cite{golubtsov}, who had proposed a closely related definition, and Cho and Jacobs~\cite{cho_jacobs}, where an equivalent definition was used under the term \emph{affine CD-category}.

@@

\section{Infinite tensor products in semicartesian monoidal categories}
\label{infprod_semicartesian}

\todo[inline]{Let's also simplify notation by writing $X_J$ and $X_F$ instead of writing $\bigotimes$ every time. This is quite standard}

Suppose $\cC$ is a semicartesian monoidal category. In our

In order to make sense of infinite families of random variables within $\cC$,
we will want to have a notion of infinite tensor product in $\cC$. By Kolmogorov's extension theorem, the joint distribution of infinitely many random variables is uniquely determined by the family of distributions of all finite subsets of these variables. This motivates a general definition of the infinite tensor product of a family of objects $(X_i)_{i \in J}$, based on the observation that if $F \subseteq F' \subseteq J$ are two finite subset, then the fact that $\cC$ is semicartesian monoidal gives us morphisms
\[
	\bigotimes_{i \in F'} X_i \longrightarrow \bigotimes_{i \in F} X_i
\]
to be interpreted as \emph{marginalization}. Via these maps, the finite tensor products $\bigotimes_{i \in F} X_i$ make up a cofiltered diagram in the form of a functor from the poset of finite subsets of $J$, ordered by reverse inclusion, to $\cC$.

@@

Of course, in a general Markov category the above cofiltered limit need not exist for infinite $J$, in which case the infinite tensor product does not exist either (as in \Cref{infprods_finstoch}).

\section{Infinite tensor products in Markov categories}
\label{infprod_markov}

@@

\todo[inline]{Perhaps we could change the terminology to ``Kolmogorov product''. I'll tentatively use this in the following}

@@

\begin{remark}
    Condition (2) is also equivalent to the statement that, given a compatible family of deterministic maps $a_F: A \to \bigotimes_{i \in F} X_i$, the induced map $a: A \to \bigotimes_i X_i$ is deterministic.
\end{remark}
